How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-09-24 analysis/paper/paper.Rmd
Able to save spd image in png

In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca. 36 ka) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421]. Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

@@

After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. This resulted in a sample of 33 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

>>>>

In Japan, Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38 ka, accompanied by remains of Homo sapiens found in Okinawa Island (ca. 36 ka) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes and led to the megafaunal extinctions [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early; @norton2010nature; @bae2017late; @nakazawa2017pleistocene; @kaifu2019palaeolithic]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. An important example is Ishinomoto 8-ku, which is one of the earliest sites, dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artifacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421]. Stemmed points appear in Japan much later than in Korea, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

@@

After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. This resulted in a sample of 33 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.
